On-the-fly Algorithms for GKAT Equivalences \thanks{Identify applicable funding agency here. If none, delete this.}

### 2025-01-07 main.tex
checkpoint symb bisim

Symbolic GKAT coalgebra, instead of computing the behavior of each atom individually, groups atoms into boolean expressions. 
This optimization leads to space-efficient coalgebras and an equivalence checking algorithm making use of off-the-shelf SAT solvers.
Specifically, given a set of primitive actions \(K\) and primitive tests \(T\), a \emph{symbolic GKAT coalgebra} \(Ŝ ≜ (S, ϵ̂, δ̂)\) consists of a state set \(S\) and an accepting function \(ϵ̂\) and a transition function \(δ̂\):
\begin{mathpar}
    ϵ̂: S → 𝒫(\BExp_T), \and
    δ̂: S → 𝒫(\BExp_T × S × K),
\end{mathpar}
where for all states \(s ∈ S\), all the boolean expressions are ``disjoint''; namely the conjunction of any two expressions from the set \(\{ϵ̂(s)\} ∪ \{b ∣ ∃ (b, s', p) ∈ δ(s)\}\) is equivalent to \(0\) under boolean algebra.

Notice that we call \(ϵ̂\) the accepting function and \(δ̂\) the transition function; intuitively, a state \(s\) accepts an atom \(α\) if and only if there exists a \(b ∈ ϵ̂(s)\), s.t. \(α ≤ b\); similarly, the state \(s\) transitions to \(s'\) via atom \(α\) while executing \(p\) if and only if there exists \((b, s', p) ∈ δ̂(s)\) and \(α ≤ b\).
With the above intuition in mind, a symbolic GKAT coalgebra \(Ŝ ≜ (S, ϵ̂, δ̂)\) can be lowered into a GKAT coalgebra \((S, δ)\) in the following manner:
\begin{align}\label{cons:lowering}
δ(s, α) ≜ \begin{cases}
    \accept & ∃ b ∈ ϵ̂(s), α ≤ b \\[5px]
    (s', p) & 
        \begin{aligned}
            & ∃ b ∈ \BExp_T, α ≤ b \\
            & \text{ and } δ(s, b) = (s', p)  
        \end{aligned}\\[5px]
    \reject & \text{otherwise}
\end{cases}
\end{align}
This is well-defined, i.e. no more than one clause can be satisfied precisely because the boolean expressions appear in \(ϵ̂\) and \(δ̂\) are disjoint.
We usually use \(S\) to denote the lowering of \(Ŝ\); and the semantics of a state \(s ∈ Ŝ\) is defined as its semantics in the lowering \(⟦s⟧_{Ŝ} ≜ ⟦s⟧_{S}.\)

We will then use \(ρ̂(s): \BExp_T\) to represent the boolean expressions that contain all the atoms that the state \(s\) rejects, and \(ρ̂(s)\) can be computed as follows:
\begin{align*}
    ρ̂(s) ≜ ⋀ \{\overline{b} ∣{}
        & ∃ s' ∈ S, p ∈ K, (b, s', p) ∈ δ̂(s) \\
        & \text{ or } b ∈ ϵ̂(s)\}.
\end{align*}
The trace semantics of a GKAT coalgebra \((S, ϵ̂, δ̂)\) is then defined as the trace semantics of its lowering \((S, δ)\).

\begin{remark}[Canonicity]\label{rem:canonicity}
    Notice that symbolic GKAT coalgebra is not canonical, i.e. there exists two different symbolic GKAT colagebra with the same lowering, consider the state set \(S ≜ \{s\}\):
    \begin{align*}
        {δ̂}₁(s) & ≜ \{b ↦ (s, p), \overline{b} ↦ (s, p)\} \\
        {δ̂}₂(s) & ≜ \{⊤ ↦ (s, p)\},
    \end{align*} 
    and both \(ϵ̂₁, ϵ̂₂\) will return constant \(0\).
    These two symbolic GKAT coalgebra \(Ŝ₁ ≜ (S, δ̂₁, ϵ̂₁)\) and \(Ŝ₂ ≜ (S, δ̂₂, ϵ̂₂)\) have the same lowering and semantics, yet, they are different.
    There are other symbolic representation that will satisfy canonicity, yet we opt to use our current representation for ease of construction and computational efficiency.
\end{remark}

\begin{theorem}[Functoriality]\label{thm:lowering-functor}
    The lowering operation is a functor, given a symbolic GKAT coalgebra homomorphism \(h: Ŝ → Û\), then \(h\) is also a homomorphism \(h: S → U\).
\end{theorem}

@@

The functoriality states that a homomorphism on two symbolic coalgebras implies a homomorphism of on their lowing; similarly, a bisimulation on symbolic GKAT coalgebra induces a bisimulation on their lowering.
However, the converse is not true, precisely because of the canonicity problem noted in~\Cref{rem:canonicity}: take the \(Ŝ₁\) and \(Ŝ₂\) in~\Cref{rem:canonicity}, because they have the same lowering, therefore the identity homomorphism is a homomorphism on their lowerings, yet there is no homomorphism from \(Ŝ₁\) to \(Ŝ₂\).

We can then migrate the normalized bisimulation algorithm to the symbolic setting, we will first prove an inductive construction theorem like~\cref{thm:recursive-construction}.

@@

\begin{proof}
    Reduces to~\cref{thm:recursive-construction} i.e. all the above condition holds if and only if all the condition in~\cref{thm:recursive-construction} holds in the lowered coalgebra.
\end{proof}

Then for the algorithm, we can just recursively check all the conditions in~\cref{thm:symb-recursive-construction}.

@@

Inspired by the syntax of Ocaml, the \(\mathrel{\&\!\&}\) is the logical-and operator on the language level, specifying that all four conditions in the return statements must be satisfied to return true. 
Notice just like the non-symbolic case, this algorithm can be modified to perform symbolic bisimulation of (non-normalized) GKAT coalgebra, which coincides with infinite trace equivalence~\cite{schmid_GuardedKleeneAlgebra_2021}, by letting \textsc{isDead} and \textsc{knownDead} to always return false.

\section{Symbolic GKAT Coalgebra Construction}

>>>>

Symbolic GKAT coalgebra, instead of computing the behavior of each atom individually, groups atoms into boolean expressions. 
This optimization leads to space-efficient coalgebras and an equivalence checking algorithm making use of off-the-shelf SAT solvers.
Specifically, given a set of primitive actions \(K\) and primitive tests \(T\), a \emph{symbolic GKAT coalgebra} \(Ŝ ≜ (S, ϵ̂, δ̂)\) consists of a state set \(S\) and an accepting function \(ϵ̂\) and a transition function \(δ̂\):
\begin{mathpar}
    ϵ̂: S → 𝒫(\BExp_T), \and
    δ̂: S → 𝒫(\BExp_T × S × K).
\end{mathpar}
This coalgebra is also required to satisfy the disjointedness condition, i.e. for all states \(s ∈ S\), the conjunction of any two expressions from the set \(ϵ̂(s) ∪ \{b ∣ ∃ (b, s', p) ∈ δ(s)\}\) is equivalent to \(0\) under boolean algebra.

We name \(ϵ̂\) the accepting function because intuitively a state \(s\) accepts an atom \(α\) when there exists a \(b ∈ ϵ̂(s)\), s.t. \(α ≤ b\); similarly, \(δ̂\) is called the transition function because a state \(s\) transitions to \(s'\) via atom \(α\) while executing \(p\) when there exists \((b, s', p) ∈ δ̂(s)\) and \(α ≤ b\).
With the above intuition in mind, a symbolic GKAT coalgebra \(Ŝ ≜ (S, ϵ̂, δ̂)\) can be lowered into a GKAT coalgebra \(S ≜ (S, δ)\) in the following manner:
\begin{align}\label{cons:lowering}
δ(s, α) ≜ \begin{cases}
    \accept & ∃ b ∈ ϵ̂(s), α ≤ b \\[5px]
    (s', p) & 
        \begin{aligned}
            & ∃ b ∈ \BExp_T, α ≤ b \\
            & \text{ and } δ(s, b) = (s', p)  
        \end{aligned}\\[5px]
    \reject & \text{otherwise}
\end{cases}
\end{align}
This is well-defined, i.e. exactly one clause can be satisfied for any \(s ∈ S\) and \(α ∈ \At\), because of the disjointedness condition.
We usually use \(S\) to denote the lowering of \(Ŝ\); and the semantics of a state \(s ∈ Ŝ\) is defined as its semantics in the lowering \(⟦s⟧_{Ŝ} ≜ ⟦s⟧_{S}.\)

We will then use \(ρ̂(s): \BExp_T\) to represent all the atoms that the state \(s\) rejects in the lowering, and \(ρ̂(s)\) is computed as follows:
\begin{align*}
    ρ̂(s) ≜ ⋀ \{\overline{b} ∣{}
        & ∃ s' ∈ S, p ∈ K, (b, s', p) ∈ δ̂(s) \\
        & \text{ or } b ∈ ϵ̂(s)\}.
\end{align*}

\begin{remark}[Canonicity]\label{rem:canonicity}
    Symbolic GKAT coalgebra is not canonical, i.e. there exists two different symbolic GKAT coalgebra with the same lowering, consider the state set \(S ≜ \{s\}\):
    \begin{align*}
        {δ̂}₁(s) & ≜ \{b ↦ (s, p), \overline{b} ↦ (s, p)\} \\
        {δ̂}₂(s) & ≜ \{⊤ ↦ (s, p)\},
    \end{align*} 
    and both \(ϵ̂₁, ϵ̂₂\) will return constant \(0\).
    These two symbolic GKAT coalgebra \(Ŝ₁ ≜ (S, δ̂₁, ϵ̂₁)\) and \(Ŝ₂ ≜ (S, δ̂₂, ϵ̂₂)\) have the same lowering and semantics, yet, they are different.
    It is possible to construct symbolic representations that satisfies canonicity, yet we opt to use our current representation for ease of construction and computational efficiency.
\end{remark}

\begin{theorem}[Functoriality]\label{thm:lowering-functor}
    The lowering operation is a functor, every symbolic GKAT coalgebra homomorphism \(h: Ŝ → Û\), is also a GKAT coalgebra homomorphism \(h: S → U\).
\end{theorem}

@@

The functoriality states that a homomorphism on two symbolic coalgebras induces a homomorphism of on their lowing; similarly, a bisimulation on symbolic GKAT coalgebra also induces a bisimulation on their lowering.
However, the converse is not true, precisely because of the canonicity problem noted in~\Cref{rem:canonicity}: take the \(Ŝ₁\) and \(Ŝ₂\) in~\Cref{rem:canonicity}, because they have the same lowering, therefore the identity homomorphism is a homomorphism on their lowerings, but there is no homomorphism from \(Ŝ₁\) to \(Ŝ₂\).

We can then define the symbolic equivalence algorithm, with its correctness stated in~\cref{thm:recursive-construction}.

@@

\begin{proof}
    Reduces to~\Cref{thm:recursive-construction} i.e. all the above condition holds if and only if all the condition in~\Cref{thm:recursive-construction} holds in the lowering.
\end{proof}

Similar to the non-symbolic version of the algorithm, we will first check if either of the input states is dead, then  recursively check all the conditions in~\cref{thm:symb-recursive-construction}.
And the symbolic version of the liveness checking algorithm is also based on a graph search algorithm on all the reachable state, where we try to identify an accepting state by checking whether \(⋁ ϵ̂(s) ≡ 0\).
The only deviation from the non-symbolic cases is that we do not follow empty transitions in the symbolic case, i.e. if \(δ̂(s) ≜ (b, s', p)\) and \(b ≡ 0\), we will not search \(s'\) to check the liveness of \(s'\).
In~\Cref{sec:optimization-implementation}, we will outline a simple modification for the symbolic GKAT coalgebra generation algorithm in~\Cref{sec:symb-gkat-construction} to avoid generating empty transitions, allowing us to skip the emptiness check when deciding whether a state is dead.

@@

The symbolic trace-equivalence algorithm is presented as~\cref{alg:symb-bisim}, where \(\mathrel{\&\!\&}\) is the logical-and operator on the language level, specifying that all the conditions in~\cref{thm:symb-recursive-construction} must be satisfied for the algorithm to return true.

\section{Symbolic GKAT Coalgebra Construction}\label{sec:symb-gkat-construction}
